iSEE: Interactive SummarizedExperiment Explorer

### 2018-05-10 manuscript/main.tex
Minor edits from John's comments.

\begin{abstract}
Data exploration is critical to the comprehension of large biological data sets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualisation are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor packages for analysing high-throughput biological data,
and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics data sets.

@@

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays commonly used in genomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualisation of the data in an intuitive and interactive interface is crucial for allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\

@@

\section*{Taking a SummarizedExperiment input}
The iSEE interface is initialised with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
Each instance of the SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
For instance, individual assays may represent gene expression matrices, either in the form of raw counts or as values obtained by different normalisation strategies.
In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'' slots, respectively; those may include experimental metadata as well as any analysis results.\\

@@

\subsection*{Using a multi-panel layout}
All data aspects stored in a SummarizedExperiment can be readily examined alongside one another in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
The interface layout is built using the shinydashboard package \citep{chang2018shinydashboard}, using colour-coded panels to visualise each data aspect.
Individual panel types include:

@@

\section*{Acknowledgements}
We thank the organisers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
We also thank members of the Bioconductor community for their helpful suggestions.

>>>>

\begin{abstract}
Data exploration is critical to the comprehension of large biological data sets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualisation are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor packages for analysing high-throughput biological data,
and provides useful features such as simultaneous examination of (meta)data and analysis results, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics data sets.

@@

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays commonly used in genomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualisation of the data using an intuitive and interactive interface is crucial for allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\

@@

\section*{Taking a SummarizedExperiment input}
The iSEE interface is initialised with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
Each instance of the SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
For instance, individual assays may represent gene expression matrices, either in the form of raw counts or as values obtained by different normalisation strategies.
In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'' slots, respectively; these may include experimental metadata as well as any analysis results.\\

@@

\subsection*{Using a multi-panel layout}
All data aspects stored in a SummarizedExperiment can be simultaneously examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
The interface layout is built using the shinydashboard package \citep{chang2018shinydashboard}, using colour-coded panels to visualise each data aspect.
Individual panel types include:

@@

\section*{Acknowledgements}
We thank the organisers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
We also thank members of the Bioconductor community for their helpful suggestions.
Finally, we thank John Marioni for helpful comments on the manuscript.
